A Horrocks-type theorem for even orthogonal $K_2$

### 2019-08-20 dl-paper.tex
Finished sketching the introduction

\title{A Horrocks-type theorem for orthogonal $K_2$}
\keywords {Steinberg group, $K_2$-functor, Serre problem
 {\em Mathematical Subject Classification (2010):} 19C20}
\author{Andrei Lavrenov}
\address{Mathematisches Institut der Universit\"at M\"unchen, Theresienstr. 39, D-80333 M\"unchen}
\email{avlavrenov at gmail.com}

@@

\begin{document}
\begin{abstract} We prove an analogue of Horrocks theorem for orthogonal Steinberg groups. \end{abstract}
\maketitle
\section{Introduction}

The solution of the famous Serre problem on projective modules by D.~Quillen and A.~Suslin was one of the achievements of the mathematics of the 70's,
 which played a significant role in the development of modern algebraic K-theory. In its simplest form it asks whether any projective module over a polynomial ring $k[x_1,\ldots, x_n]$ over a field $k$ is free. We refer the reader to the textbook~\cite{Lam10} for a more complete introduction to the problem and its history.

Numerous analogues of Serre problem have been studied subsequently. For example, A.~Suslin himself solved the so-called $K_1$-analogue of Serre's problem, see~\cite{Su77}.
His results were generalized by E.~Abe in~\cite{Abe83} and later by A.~Stavrova in~\cite{St-poly}.

There are three main ingredients in the proof of the $\K_2$-analogue of Serre's problem. These ingredients are Quillen--Suslin local-global principle, Horrocks theorem, Zariski glueing.
In~\cite{LS17} the authors have already shown that Steinberg groups $\St(\Phi, R)$ satisfy the local-global principle. In this paper we make another step towards
 transferring Tulenbayev's result to orthogonal Steiberg groups and prove the following theorem (cf.~\cite[Theorem~5.1]{Tu83})
\begin{theorem}\label{thm:main} Suppose $A$ is a commutative ring in which $2$ is invertible. Then for any $\ell \geq 7$ the following commutative square is a pullback in which all arrows are injective:
\begin{equation} \nonumber
  \xymatrix{ \St(\rD_\ell, A) \ar[r] \ar[d] & \St(\rD_\ell, A[X]) \ar[d] \\ \St(\rD_\ell, A[X\inv]) \ar[r] & \St(\rD_\ell, A[X, X\inv]).}
\end{equation} \end{theorem}
It is easy to see that in the statement of the above theorem the groups $\St(\rD_\ell, -)$  can be replaced with the groups $\K_2(\rD_\ell, -)$. Thus, the above result can be thought of as a direct $\K_2$-analogue of~\cite[Theorem~6.8]{Su82}.

The proof of~\cref{thm:main} goes as follows. After invoking the local-global principle~\cite[Theorem~2]{LS17} the proof reduces to the special case when $A$ is local. Now if $M$ is the maximal ideal of $A$, the proof of the injectivity of $\St(\rD_\ell, A[X]) \to \St(\rD_\ell, A[X, X\inv])$ comes down to proving injectivity of the following two maps
\[ \xymatrix{ \St(\rD_\ell, A[X]) \ar[r] & \St(\rD_\ell, A[X\inv] + M[X]) \ar[r] & \St(\rD_\ell, A[X, X\inv].) }\]
The injectivity of the second map is obtained in~\cref{thm41}. This is the only place in our proof which uses the invertibility of $2$ assumption. This result involves some previously known computations with Grothendieck--Witt groups and also the stabilization theorem for the orthogonal K-theory. The injectivity of the first map is obtained in~\cref{thm:P1glueing}, which is a direct generalization of~\cite[Proposition~4.3]{Tu83}. In the proof of this result we also make explicit many of the details which are left implicit in Tulenbaev's proof.

\subsection{Acknowledgements} Theorem 2 is due to the first-named author who acknowledges the financial support of RSCF grant No ???.
Theorem 3 is due to the second-named author who acknowledges the financial support of JSC Gazpromneft and RFBR grant No 18-31-20044

The authors would also to express their gratitude to A.~Stavrova for posing the problem and also to A.~Stepanov and N.~Vavilov for many useful comments and also their interest in this work.

@@

\subsection{Steinberg groups}
Let $\Phi$ be a reduced and irreducible root system of rank $\geq 2$ and $R$ be a commutative ring with $1$. Recall that in this case the \emph{Steinberg group} $\St(\Phi, R)$ can be defined by means of generators $x_{\alpha}(s)$ 
%(where $\alpha\in\Phi, s\in R$)
and relations:
\begin{align}
& \phantom{[}
x_\alpha(s) \cdot x_\alpha(t) = x_\alpha(s+t),\ \alpha\in\Phi,\ s,t\in R; \label{rel:add}\\
& [x_\alpha(s), x_\beta(t)] = \prod\limits_{i,j\in\mathbb{N}}
 x_{i\alpha + j\beta}\left(N_{\alpha,\beta ij}\, s^i t^j\right),\quad \alpha,\beta\in\Phi,\ \alpha\neq\pm\beta,\ s,t\in R. \label{rel:CCF}
\end{align}
The indices $i$, $j$ appearing in the right-hand side of the above relation range over
all positive natural numbers such that $i\alpha + j\beta\in\Phi$.
The constants $N_{\alpha \beta i j}$ appearing in the right-hand side of \eqref{rel:CCF} are integers equal to $\pm 1,\pm 2,\pm 3$, they are called the {\it structure constants} of the Chevalley group $G(\Phi, R)$. Several different methods of computing signs of these constants have been proposed in the literature, see e.\,g.~\cite{V00}, \cite[\S~9]{VP}.

@@

 Now if $g \in \St(\rD_\ell, A[X])$ is such that its image in $\St(\rD_\ell, A[X, X\inv])$ is trivial, then so is its image in all localizations $\St(\rD_\ell, A_M[X, X\inv])$, where $M$ ranges over the maximal ideals of $A$. By the just established local case the images of $g$ in all $\St(\rD_\ell, A_M[X])$ are also trivial hence by the local-global principle \cite[Theorem~2]{LS17} the element $g$ is also trivial.

>>>>

\title{A Horrocks-type theorem for even orthogonal $K_2$}
\keywords {Steinberg group, $K_2$-functor, Serre problem
 {\em Mathematical Subject Classification (2010):} 19C20}
\author{Andrei Lavrenov}
\address{Mathematisches Institut der Universit\"at M\"unchen, Theresienstr. 39, D-80333 M\"unchen}
\email{avlavrenov at gmail.com}

@@

\begin{document}
\begin{abstract} We prove an analogue of Horrocks theorem for even-dimensional orthogonal Steinberg groups. This result transfers to the orthogonal case an earlier result of M.~Tulenbaev and is also a $\K_2$-analogue of an earlier result of A.~Suslin and V.~Kopeiko. \end{abstract}
\maketitle
\section{Introduction}
The solution of the famous Serre problem on projective modules by D.~Quillen and A.~Suslin was one of the brilliant achievements of mathematics of the 70's, which bolstered the development of modern algebraic K-theory. In its weakest form Serre problem asks whether any projective module over a polynomial ring $R = k[x_1,\ldots, x_n]$ over a field $k$ is free. We refer the reader to the textbook~\cite{Lam10} for a comprehensive account on the problem, its history and the subsequent affirmative solution.

After the original Serre problem had been solved, numerous analogous statements have drawn the attention of specialists. For example, A.~Suslin formulated and solved the so-called $K_1$-analogue of Serre's problem. This result asserts that the functor $\K_{1,n}(R) = \mathrm{GL}_{n}(R)/\E_n(R)$ has trivial values on polynomial rings $k[x_1, \ldots x_n]$ over fields for all $n \geq 3$ see~\cite[Corollary~7.11]{Su77}. Suslin's results were subsequently generalized to $\K_1$-functors modeled on other linear groups (see the definition below). For example, for even-dimensional orthogonal groups the corresponding result was obtained by A.~Suslin and V. Kopeiko in~\cite{Su82}, while for more general types of Chevalley groups of rank $\geq 2$ this is a result of E.~Abe, see~\cite{Abe83}. Recently, A.~Stavrova has obtained probably the most general result in this direction: she solved the analogue of Serre problem for the functor $\K_1^G$ modeled on arbitary isotropic reductive group scheme $G/k$ of isotropic rank $\geq 2$ (see~\cite[Theorem~1.2]{St-poly}).

Recall that to every irreducible root system $\Phi$ and a commutative ring $R$ one can associate two groups: the group $G(\Phi, R)$ called {\it simply-connected Chevalley group} (see~\cite[\S~3]{St71} or~\cite{VP}) and the group $\St(\Phi, R)$ called {\it Steinberg group} (see~\cref{sec:Steinberg-intro} for the definition). There is a well-defined map $\pi \colon \St(\Phi, R) \to G(\Phi, R)$ sending each generator $x_\alpha(\xi)$ to the elementary root unipotent $t_\alpha(\xi)$. The cokernel and the kernel of this map are, by definition, the $\K_1$ and $\K_2$ groups modeled on Chevalley group $G(\Phi, -)$:
\[ \xymatrix{ 1 \ar[r] & \K_2(\Phi, R) \ar[r] & \St(\Phi, R) \ar^{\pi}[r] & G(\Phi, R) \ar[r] & \K_1(\Phi, R) \ar[r] & 1.} \]

One might ask whether any analogue of Serre's problem also exists for the functor $\K_2$ and it turns out that it is indeed so. More precisely, in~\cite{Tu83} M. Tulenbaev demonstrates an ``early stability theorem'' from which the isomorphism $\K_2(\rA_\ell, k[x_1, \ldots x_n]) \cong \K_2(\rA_\ell, k) = \K^M_2(k)$ immediately follows for $\ell \geq 4$. While a number of new results on the $\K_1$-analogue of Serre's problem have appeared in the literature since~\cite{Su77} (see e.\,g.~\cite{Abe83, St-poly}), little progress has been made for the $\K_2$-analogue since~\cite{Tu83}. Currently the conjecture still remains open for all $\Phi$ different from $\rA_\ell$, $\ell \geq 4$.

In~\cite{LS17} the authors have shown that Steinberg groups $\St(\Phi, R)$ satisfy Quillen--Suslin local-global principle provided $\Phi$ is simply-laced and has rank $\geq 3$. This result is one of the ingredients needed in the proof of the $\K_2$-analogue of Serre problem. The aim of the present article is to make yet another step towards the solution of the problem. Our main result is the following theorem, which is the orthogonal analogue of~\cite[Theorem~5.1]{Tu83} and the $\K_2$-analogue of~\cite[Theorem~6.8]{Su82} (cf. also with~\cite[Theorem~VI.5.2]{Lam10} and~\cite[Theorem~1.1]{St-poly}).
\begin{theorem}[Horrocks theorem for orthogonal $\K_2$]\label{thm:main} Suppose $A$ is a commutative ring in which $2$ is invertible. Then for any $\ell \geq 7$ the following commutative square is a pullback square in which all arrows are injective:
\begin{equation} \nonumber
  \xymatrix{ \St(\rD_\ell, A) \ar[r] \ar[d] & \St(\rD_\ell, A[X]) \ar[d] \\ \St(\rD_\ell, A[X\inv]) \ar[r] & \St(\rD_\ell, A[X, X\inv]).}  
\end{equation} 
Moreover, the same assertion remains true if one replaces the functor $\St(\rD_\ell, -)$ with $\K_2(\rD_\ell, -)$. \end{theorem}

Our proof of~\cref{thm:main} goes as follows. After invoking the local-global principle~\cite[Theorem~2]{LS17} the proof reduces to the special case when $A$ is local. Now if $M$ is the maximal ideal of $A$, the proof of the injectivity of $\St(\rD_\ell, A[X]) \to \St(\rD_\ell, A[X, X\inv])$ comes down to proving injectivity of the following two maps
\[ \xymatrix{ \St(\rD_\ell, A[X]) \ar[r] & \St(\rD_\ell, A[X\inv] + M[X]) \ar[r] & \St(\rD_\ell, A[X, X\inv].) }\]
The injectivity of the second map is obtained in~\cref{thm41}. This is the only place in our proof which invokes the assumption that $2$ is invertible. The proof of~\cref{thm41} depends on certain computations with Grothendieck--Witt groups and also on the stability theorem for the higher orthogonal K-groups. The injectivity of the first map is obtained in~\cref{thm:P1glueing}, which is a direct generalization of~\cite[Proposition~4.3]{Tu83}. %In the proof of~\cref{thm:P1glueing} we also write explicitly many of the details left implicit in Tulenbaev's paper.

\subsection{Acknowledgements} Theorem 2 is obtained by the first-named author who acknowledges the financial support of RSCF grant No ???.
Theorem 3 is obtained by the second-named author who acknowledges the financial support of JSC Gazpromneft and RFBR grant No 18-31-20044

The authors would also to express their gratitude to A.~Stavrova for posing the problem and also to A.~Stepanov and N.~Vavilov for many useful comments and their interest in this work.

@@

\subsection{Steinberg groups} \label{sec:Steinberg-intro}
Let $\Phi$ be a reduced and irreducible root system of rank $\geq 2$ and $R$ be a commutative ring with $1$. Recall that in this case the \emph{Steinberg group} $\St(\Phi, R)$ can be defined by means of generators $x_{\alpha}(s)$ 
%(where $\alpha\in\Phi, s\in R$)
and relations:
\begin{align}
& \phantom{[}
x_\alpha(s) \cdot x_\alpha(t) = x_\alpha(s+t),\ \alpha\in\Phi,\ s,t\in R; \label{rel:add}\\
& [x_\alpha(s), x_\beta(t)] = \prod\limits_{i,j\in\mathbb{N}}
 x_{i\alpha + j\beta}\left(N_{\alpha,\beta ij}\, s^i t^j\right),\quad \alpha,\beta\in\Phi,\ \alpha\neq\pm\beta,\ s,t\in R. \label{rel:CCF}
\end{align}
The indices $i$, $j$ appearing in the right-hand side of the above relation range over
all positive natural numbers such that $i\alpha + j\beta\in\Phi$.
The constants $N_{\alpha \beta i j}$ appearing in the right-hand side of \eqref{rel:CCF} are integers equal to $\pm 1,\pm 2,\pm 3$, they are called the {\it structure constants} of the Chevalley group $G(\Phi, R)$. Several different methods of computing signs of these constants have been proposed in the literature, see e.\,g.~\cite{V00}, \cite[\S~9]{VP}.

@@

 Now if $g \in \St(\rD_\ell, A[X])$ is such that its image in $\St(\rD_\ell, A[X, X\inv])$ is trivial, then so is its image in all localizations $\St(\rD_\ell, A_M[X, X\inv])$, where $M$ ranges over the maximal ideals of $A$. By the just established local case the images of $g$ in all $\St(\rD_\ell, A_M[X])$ are also trivial. Now by the local-global principle \cite[Theorem~2]{LS17} the element $g$ is trivial as well.
